---
title: "*B. distachyon - H. vulgare*"
author: Root samples
output:
  html_document:
    theme: yeti
    code_folding: hide
    number_sections: false
editor_options: 
  chunk_output_type: console
---

```{r libraries, include=FALSE}
# Core libraries
library(tidyverse)

# Global options
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)

# Plot settings
theme_set(theme_classic())
theme_update(plot.title = element_text(face = "bold"))

# Additional libraries
required_packages <- c("gplots", "DT", "RColorBrewer", "cowplot", "matrixStats", "readr")
if (!all(required_packages %in% installed.packages())) {
  install.packages(setdiff(required_packages, rownames(installed.packages())))
}

lapply(required_packages, library, character.only = TRUE)
```


### Set up species and input files

```{r setup}

# Species names
species1_name <- "Brachypodium_distachyon"
species2_name <- "Hordeum_vulgare"

# Expression files
species1_expr_file <- "~/Data/BDIS_R_vst-normalized.tsv"
species2_expr_file <- "~/Data/HVUL_R_vst-normalized.tsv"

# Ortholog file
ortholog_group_file <- "~/HOG/Brachypodium_distachyon__v__Hordeum_vulgare.tsv"

# Gene version patterns 
species1_gene_version_number <- ".v3.2"
species2_gene_version_number <- ".V3"
ortholog_group_file_gene_version_numbers <- "\\.[0-9]{1,2}\\.[a-zA-Z]$|\\.[0-9]{1,2}$"

# Comparison table object name
comparison_table_name <- "bdis_hvul_r"
```



### Read in ortholog groups and expression data

```{r input, warning=FALSE, error=FALSE, message=FALSE}

# Read in ortholog groups
# =======================

ortholog_group_RData <- paste0("orthologs-", species1_name, "-", species2_name, ".RData")

if (!file.exists(ortholog_group_RData)) {

  ortho <- read_tsv(ortholog_group_file, show_col_types = FALSE) %>%
    dplyr::rename(Species1 = species1_name, Species2 = species2_name) %>%
    dplyr::select(Species1, Species2, Orthogroup) %>%
    filter(Species1 != "", Species2 != "") %>%
    separate_rows(Species1, sep = ", ", convert = FALSE) %>%
    separate_rows(Species2, sep = ", ", convert = FALSE) %>%
    group_by(Species1, Species2, Orthogroup) %>%
    dplyr::slice(1) %>% 
    ungroup() 

  save(ortho, file = ortholog_group_RData)
} else {
  load(file = ortholog_group_RData)
}

# Read in expression data
# =======================

species1_expr <- read.delim(species1_expr_file, sep = "\t", header = TRUE)
species2_expr <- read.delim(species2_expr_file, sep = "\t", header = TRUE)

# Remove gene version numbers
# ======================

if (!is.null(ortholog_group_file_gene_version_numbers)) {
  ortho <- ortho %>% 
    mutate(Species1 = str_remove_all(Species1, ortholog_group_file_gene_version_numbers)) %>% 
    mutate(Species2 = str_remove_all(Species2, ortholog_group_file_gene_version_numbers))
}

species1_expr <- species1_expr %>%
  mutate(GeneID = if (!is.null(species1_gene_version_number)) {
    str_remove(GeneID, species1_gene_version_number)
  } else {
    GeneID
  }) %>%
  dplyr::rename(Genes = GeneID)

species2_expr <- species2_expr %>%
  mutate(GeneID = if (!is.null(species2_gene_version_number)) {
    str_remove(GeneID, species2_gene_version_number)
  } else {
    GeneID
  }) %>%
  dplyr::rename(Genes = GeneID)


# Filter
# ======

cat (length(unique(ortho$Orthogroup)), " ortholog groups containing:\n",
     " ", length(unique(ortho$Species1)), " ", species1_name, " genes\n", 
     " ", length(unique(ortho$Species2)), " ", species2_name, " genes\n\n",
     length(unique(species1_expr$Genes)), " expressed ", species1_name, " genes\n",
     length(unique(species2_expr$Genes)), " expressed ", species2_name, " genes\n",
     sep = "") 

ortho <- ortho %>%
  filter(Species1 %in% species1_expr$Genes & Species2 %in% species2_expr$Genes)

species1_expr <- species1_expr[species1_expr$Genes %in% ortho$Species1,]
species2_expr <- species2_expr[species2_expr$Genes %in% ortho$Species2,]

cat ("After filtering on expressed genes with ortholog:\n",
     "  ", length(unique(ortho$Orthogroup)), " ortholog groups containing: \n",
     "  ", length(unique(ortho$Species1)), " ", species1_name, " genes\n",
     "  ", length(unique(ortho$Species2)), " ", species2_name, " genes\n",
     sep = "")

```


### Compute co-expression networks

Several parameters:

- Correlation: Pearson or Spearman
- Normalization: CLR or MR
- Network density threshold -> Co-expression threshold

```{r coexpression}
cor_method <- "pearson" # pearson spearman
cor_sign <- "" # abs
norm_method <- "MR" # CLR MR
density_thr <- 0.03
randomize <- "" # rand

comparison_RData <- paste0("comparison-", species1_name, "-", species2_name, "-", 
                           cor_sign, cor_method, norm_method, density_thr, randomize, ".RData")

if (!file.exists(comparison_RData)) {
  
  if (randomize == "rand") {
    species1_expr$Genes <- sample(species1_expr$Genes, nrow(species1_expr), FALSE)
    species2_expr$Genes <- sample(species2_expr$Genes, nrow(species2_expr), FALSE)
  }
  
  species1_net <- cor(t(species1_expr[,-1]), method = cor_method)
  dimnames(species1_net) <- list(species1_expr$Genes, species1_expr$Genes)
  
  species2_net <- cor(t(species2_expr[,-1]), method = cor_method)
  dimnames(species2_net) <- list(species2_expr$Genes, species2_expr$Genes)
  
  if (cor_sign == "abs") {
    species1_net <- abs(species1_net)
    species2_net <- abs(species2_net)
  }
  
  if (norm_method == "CLR") {
    
    z <- scale(species1_net)
    z[z < 0] <- 0
    species1_net <- sqrt(t(z)**2 + z**2)
    
    z <- scale(species2_net)
    z[z < 0] <- 0
    species2_net <- sqrt(t(z)**2 + z**2)
    
  } else if (norm_method == "MR") {
    R <- t(apply(species1_net, 1, rank))
    species1_net <- sqrt(R * t(R))
    
    R <- t(apply(species2_net, 1, rank))
    species2_net <- sqrt(R * t(R))
  }
  
  diag(species1_net) <- 0
  diag(species2_net) <- 0
  
  R <- sort(species1_net[upper.tri(species1_net, diag = FALSE)], decreasing = TRUE)
  species1_thr <- R[round(density_thr*length(R))]
  plot(density(R), xlab = paste0(species1_name, " correlations"), main = "")
  
  R <- sort(species2_net[upper.tri(species2_net, diag = FALSE)], decreasing = TRUE)
  species2_thr <- R[round(density_thr*length(R))]
  plot(density(R), xlab = paste0(species2_name, " correlations"), main = "")
  
} else {
  load(file = comparison_RData)
}

cat(species1_name, "co-expr threshold", format(species1_thr, digits = 3) , "\n")
cat(species2_name, "co-expr threshold", format(species2_thr, digits = 3) , "\n")

```

## Network comparison

For each ortholog pair, p-values for for the overlap of network neighborhoods are computed in both direction. 

```{r comparison, warning=FALSE, error=FALSE, message=FALSE}

if (!file.exists(comparison_RData)) {

  comparison <- ortho
  
  comparison$Species1.neigh <- c(NA)
  comparison$Species1.ortho.neigh <- c(NA)
  comparison$Species1.neigh.overlap <- c(NA)
  comparison$Species1.p.val <- c(NA)
  
  comparison$Species2.neigh <- c(NA)
  comparison$Species2.ortho.neigh <- c(NA)
  comparison$Species2.neigh.overlap <- c(NA)
  comparison$Species2.p.val <- c(NA)

  for (i in 1:nrow(ortho)) {
    
    if (i %% 100 == 0) {
      cat(i, "\n")
    }
    
    # Species 1 -> Species 2
    
    neigh <- species1_net[ortho$Species1[i],]
    neigh <- names(neigh[neigh >= species1_thr])
    
    ortho_neigh <- species2_net[ortho$Species2[i],]
    ortho_neigh <- names(ortho_neigh[ortho_neigh >= species2_thr])
    ortho_neigh <- unique(ortho$Species1[ortho$Species2 %in% ortho_neigh])
    
    N <- nrow(species1_expr)
    m <- length(neigh)
    n <- N-m
    k <- length(ortho_neigh)
    x <- length(intersect(neigh, ortho_neigh))
    p_val <- 1
    if (x > 1) {
      p_val <- phyper(x-1, m, n, k, lower.tail = FALSE)
    }
    
    comparison$Species1.neigh[i] <- m
    comparison$Species1.ortho.neigh[i] <- k
    comparison$Species1.neigh.overlap[i] <- x
    comparison$Species1.p.val[i] <- p_val
    
    # Species 2 -> Species 1
    
    neigh <- species2_net[ortho$Species2[i],]
    neigh <- names(neigh[neigh >= species2_thr])
    
    ortho_neigh <- species1_net[ortho$Species1[i],]
    ortho_neigh <- names(ortho_neigh[ortho_neigh >= species1_thr])
    ortho_neigh <- unique(ortho$Species2[ortho$Species1 %in% ortho_neigh])
    
    N <- nrow(species2_expr)
    m <- length(neigh)
    n <- N-m
    k <- length(ortho_neigh)
    x <- length(intersect(neigh, ortho_neigh))
    p_val <- 1
    if (x > 1) {
      p_val <- phyper(x-1, m, n, k, lower.tail = FALSE)
    }
    
    comparison$Species2.neigh[i] <- m
    comparison$Species2.ortho.neigh[i] <- k
    comparison$Species2.neigh.overlap[i] <- x
    comparison$Species2.p.val[i] <- p_val
  }
  
  save(comparison, species1_thr, species2_thr, file = comparison_RData)
}

# Filter orthologs not in the networks
comparison <- comparison %>%
  filter(Species1.neigh.overlap > 0 & Species2.neigh.overlap > 0)

# FDR correction
comparison$Species1.p.val <- p.adjust(comparison$Species1.p.val, method = "fdr")
comparison$Species2.p.val <- p.adjust(comparison$Species2.p.val, method = "fdr")

cat ("After filtering on gene pairs in the networks:\n",
     "  ", length(unique(comparison$Orthogroup)), " ortholog groups containing: \n",
     "  ", length(unique(comparison$Species1)), " ", species1_name, " genes\n",
     "  ", length(unique(comparison$Species2)), " ", species2_name, " genes\n",
     sep = "")

# Comparsion of p-values of orthologs: species 1 -> species 2 vs species 2 -> species 1
R <- cor.test(-log10(comparison$Species1.p.val), -log10(comparison$Species2.p.val))

data.frame(s1 = -log10(comparison$Species1.p.val),
           s2 = -log10(comparison$Species2.p.val)) %>%
  ggplot(aes(x = s1, y = s2)) +
  xlab(paste0(species1_name, " p-value (-log10)")) +
  ylab(paste0(species2_name, " p-value (-log10)")) +
  geom_point() + 
  geom_smooth(method=lm, formula = y ~ x, fill = "gainsboro") +
  ggtitle(paste0("Correlation = ", format(R$estimate, digits = 3)))


# Print some summary statistics
# =============================

# Gene pairs

cat("GENE PAIRS:\n",
    sum(comparison$Species1.p.val< 0.05), " conserved ", species1_name, " gene pairs (", 
    format(sum(comparison$Species1.p.val< 0.05)/nrow(comparison), digits = 3), ")\n", 
    sum(comparison$Species2.p.val< 0.05), " conserved ", species2_name, " gene pairs (", 
    format(sum(comparison$Species2.p.val< 0.05)/nrow(comparison), digits = 3), ")\n",
    sum(comparison$Species1.p.val< 0.05 & comparison$Species2.p.val< 0.05), " reciprocally conserved gene pairs (", 
    format(sum(comparison$Species1.p.val< 0.05 & comparison$Species2.p.val< 0.05)/nrow(comparison), digits = 3), ")\n", sep = "")

# Genes

comparison_species1 <- comparison %>%
  group_by(Species1) %>%
  arrange(Species1.p.val) %>%
  slice(1)

comparison_species2 <- comparison %>%
  group_by(Species2) %>%
  arrange(Species2.p.val) %>%
  slice(1)

comparison_species1_12 <- comparison %>%
  rowwise() %>%
  mutate(Max.p.val = max(Species1.p.val, Species2.p.val)) %>%
  group_by(Species1) %>%
  arrange(Max.p.val) %>%
  slice(1)

comparison_species2_12 <- comparison %>%
  rowwise() %>%
  mutate(Max.p.val = max(Species1.p.val, Species2.p.val)) %>%
  group_by(Species2) %>%
  arrange(Max.p.val) %>%
  slice(1)

cat("GENES:\n",
    sum(comparison_species1$Species1.p.val< 0.05), " conserved ", species1_name, " genes (", 
    format(sum(comparison_species1$Species1.p.val< 0.05)/nrow(comparison_species1), digits = 3), ")\n", 
    sum(comparison_species2$Species2.p.val< 0.05), " conserved ", species2_name, " genes (", 
    format(sum(comparison_species2$Species2.p.val< 0.05)/nrow(comparison_species2), digits = 3), ")\n",
    sum(comparison_species1_12$Max.p.val< 0.05), " reciprocally conserved ", species1_name, " genes (", 
    format(sum(comparison_species1_12$Max.p.val< 0.05 & 
               comparison_species1_12$Max.p.val< 0.05)/nrow(comparison_species1_12), digits = 3), ")\n",
    sum(comparison_species2_12$Max.p.val< 0.05), " reciprocally conserved ", species2_name, " genes (", 
    format(sum(comparison_species2_12$Max.p.val< 0.05 & 
               comparison_species2_12$Max.p.val< 0.05)/nrow(comparison_species2_12), digits = 3), ")\n",
    sep = "")

# Orthogroups

comparison_species1 <- comparison %>%
  group_by(Orthogroup) %>%
  arrange(Species1.p.val) %>%
  slice(1)

comparison_species2 <- comparison %>%
  group_by(Orthogroup) %>%
  arrange(Species2.p.val) %>%
  slice(1)

comparison_species12 <- comparison %>%
  rowwise() %>%
  mutate(Max.p.val = max(Species1.p.val, Species2.p.val)) %>%
  group_by(Orthogroup) %>%
  arrange(Max.p.val) %>%
  slice(1)

cat("ORTHOLOG GROUPS:\n",
    sum(comparison_species1$Species1.p.val< 0.05), " conserved ", species1_name, " Orthogroups (", 
    format(sum(comparison_species1$Species1.p.val< 0.05)/nrow(comparison_species1), digits = 3), ")\n", 
    sum(comparison_species2$Species2.p.val< 0.05), " conserved ", species2_name, " Orthogroups (", 
    format(sum(comparison_species2$Species2.p.val< 0.05)/nrow(comparison_species2), digits = 3), ")\n",
    sum(comparison_species12$Max.p.val< 0.05), " reciprocally conserved Orthogroups (", 
    format(sum(comparison_species12$Max.p.val< 0.05 & 
               comparison_species12$Max.p.val< 0.05)/nrow(comparison_species12), digits = 3), ")\n", sep = "")

```

Table of all comparisons. 

```{r, genepairs, warning=FALSE, error=FALSE, message=FALSE}

comparison_table <- comparison %>%
   rowwise() %>%
   mutate(Max.p.val = max(Species1.p.val, Species2.p.val)) %>%
   select(-c("Species1.neigh", "Species1.ortho.neigh", "Species2.neigh", "Species2.ortho.neigh")) %>%
   arrange(Max.p.val)

assign(comparison_table_name, comparison_table)
save(list = comparison_table_name, file = paste0(comparison_table_name, "_tbl.RData"))

comparison_table$Species1.p.val <- format(comparison_table$Species1.p.val, digits = 3, scientific = TRUE)
comparison_table$Species2.p.val <- format(comparison_table$Species2.p.val, digits = 3, scientific = TRUE)
comparison_table$Max.p.val <- format(comparison_table$Max.p.val, digits = 3, scientific = TRUE)

datatable(comparison_table, rownames = FALSE, filter = "top",
          options = list(
            columnDefs = list(list(className = 'dt-center', targets = "_all"))
            )
          )

```